---
title: "Notes_hw4"
author: "Justin Sarna"
date: "April 11, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
```{r}
```{r}
# make predictions using test data
logmod1_pred <- predict(logmod1, newdata=acs_test, type ="response" )
yhat_logmod1 <- (logmod1_pred > 0.5) * 1
head(cbind(yhat_logmod1,logmod1_pred))

logmod2_pred <- predict(logmod2, acs_test, type="response" ) 

logmod3_pred <- predict(logmod3, acs_test, type="response" ) 
logmod4_pred <- predict(logmod4, acs_test, type="response" )
logmod5_pred <- predict(logmod4, acs_test, type="response" )
linear2_pred <- predict(linear_mod2, acs_test, type="response")

# Confusion matrices
#confusionMatrix(as.factor(yhat_logmod5), as.factor(acs_fit$HighIncome), positive = "1")
log_cm1_pred <- confusionMatrix((logmod1_pred), acs_test$HighIncome, positive = "1")
log_cm2_pred <- confusionMatrix(logmod2_pred, acs_test$HighIncome, positive = "1")
log_cm3_pred <- confusionMatrix(logmod3_pred, acs_test$HighIncome, positive = "1")
log_cm4_pred <- confusionMatrix(logmod4_pred, acs_test$HighIncome, positive = "1")
log_cm5_pred <- confusionMatrix(logmod5_pred, acs_test$HighIncome, positive = "1")
linear_cm2_pred <- confusionMatrix(linear2_pred, acs_test$HighIncome, positive = "1")

# Display comparison of accuracy of each decision tree 
sprintf("Tree1: Fit acc = %.3f Pred acc = %.3f",log_cm1$overall['Accuracy'], log_cm1_pred$overall['Accuracy'])
sprintf("Tree2: Fit acc = %.3f Pred acc = %.3f",log_cm2$overall['Accuracy'], log_cm2_pred$overall['Accuracy'])
sprintf("Tree3: Fit acc = %.3f Pred acc = %.3f",log_cm3_cm$overall['Accuracy'], log_cm3_pred$overall['Accuracy'])
sprintf("Tree4: Fit acc = %.3f Pred acc = %.3f",log_cm4_cm$overall['Accuracy'], log_cm4_pred$overall['Accuracy'])
sprintf("Tree4: Fit acc = %.3f Pred acc = %.3f",log_cm5_cm$overall['Accuracy'], log_cm5_pred$overall['Accuracy'])
sprintf("Tree4: Fit acc = %.3f Pred acc = %.3f",linear_cm2$overall['Accuracy'], linear_cm2_pred$overall['Accuracy'])
```
```

